Learning Causation from Data

### 2016-03-22 talks/CausalBN/cbn2012.tex
x

\documentclass{beamer}
\usepackage{amsmath,amsfonts}	% use math symbols
\usepackage{graphicx} % insert images
\usepackage{url} % insert urls
\usepackage{textpos} % package for the positioning
\usetheme{Ilmenau} % change theme

\begin{document}

%% MAKE TITLE
\title{Learning from Causation}
\subtitle{Fundamental of Casual Inference and its Applications}
\author{Huang Xiao} %\\ \small{xiaohu@in.tum.de} }%\thanks{Funded by ARAMiS project}}
%\and Author2 \inst{1} }
%\and Author3 \inst{2}}
\institute[Technische Universit\"at M\"unchen]
{ 
	%\inst{1} 
	Chair of IT Security (I20) \\Department of Informatics \\Technische Universit\"at M\"unchen 
	%\and
	%\inst{2} Cloud Service Lab\\Fraunhofer AISEC
}

@@

%% START PRESENTATION
\section{Fundamental of Causal Inference}
\subsection{Motivation}
\begin{frame}{What is Causality?}
\begin{block}{A definition from Wikipedia}
\textbf{Causality} (also referred to as causation) is the relationship
between an event (\textit{the cause}) and a second event (\textit{the effect}),
where the second event is understood as a consequence of the
first.
\end{block}\pause
\textbf{An example in real life} : Does smoking cause lung cancer?\\
\pause	
\begin{center}
\alert{Yes, it might be!}
\end{center}
\end{frame}
\begin{frame}{From Probabilistic View }
\textbf{Problem:} Does smoking cause lung cancer?\pause
\begin{center}
\begin{itemize}
\item Smoking does \alert{increase the probability} of getting lung cancer.
\end{itemize}
\end{center}
\end{frame}
\begin{frame}{Statistical Inference Overview}
\begin{figure}
\includegraphics[scale=0.7]{imgs/statInf}
\end{figure}
\begin{itemize}
\item Approximate an estimate of $X$ given evidence $e$, namely, $Pr(X\,|\,e)$. E.g., Regression or Classification problems.
\item Rejection of hypothesis, i.e., assert wether samples are from a certain distribution.  
\item Confidence interval, i.e., construct an interval based on dataset 
\end{itemize}
\end{frame}
\begin{frame}{Causal Inference Overview}
\begin{figure}
\includegraphics[scale=0.6]{imgs/causalInf}
\end{figure}
\begin{itemize}
\item What if $P$ has shifted itself to $P^*$?
\item<2-> \textbf{Key factors:} Causes, Changes, and Invariants . 
\item<2-> Inference of $P^*$ and reasoning of changes. 
\end{itemize}
\end{frame}
\begin{frame}{What makes Causal Inference interesting?}
\begin{itemize}
\item Human understands the world in terms of causes and effects.
\item Empirical science is about establishing causes.
\item Causal inference gives a mathematical language for causal
statements, and tools to solve causal problems formally.
\item Alternative exercising to decision making, reasoning, etc.
\end{itemize}
%\begin{alertblock}{Note}
%Causal Inference has fundamental difference with machine learning
%\end{alertblock}
\end{frame}
\subsection{Causal Graphical Model}
\begin{frame}{Association}
\begin{itemize}
\item Now we want to find out what \alert{causes} lung cancer
\end{itemize} \pause
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{table}
\centering
\begin{tabular}{|c|c|c|c|}
\hline
& & \multicolumn{2}{|c|}{Lung cancer}\\\hline
smoking & yellow teeth & yes & no\\\hline
yes & yes & 100 & 400\\\hline
yes & no & 100 & 400\\\hline
no & yes & 1 & 450\\\hline
no & no & 9 & 8540\\\hline
\end{tabular}
\caption{Data observations from 10000 people}
\end{table}
\end{column}\hfill
\begin{column}{0.5\textwidth}
\vspace*{-0.5in}
\begin{figure}
\centering
\includegraphics[scale=0.6]{imgs/smokepl}
\end{figure}
\end{column}
\end{columns}
\end{frame}
\begin{frame}{Measurements of Association}
\textbf{To find out associations among variables}
\begin{itemize}
\item Mutual information (Information theory)
\item Pearson (linear) correlation
\item Spearman's rho (rank correlation)
\item Effect size between two variables
\item Many others
\end{itemize}
\end{frame}
\begin{frame}{Observations from Data}
\textbf{Obviously}
\begin{itemize}
\item \textit{yellow teeth} and \textit{lung cancer} are associated.
\end{itemize}\pause
\alert{\textbf{But...}}
\begin{itemize}
\item Bleaching the teeth does not help reduce the probability
of getting lung cancer.
\end{itemize}\pause
\begin{alertblock}{Caution!}
Correlation does not imply Causation
\end{alertblock}
\end{frame}
%% START NEW SECTION
\section{History}

\begin{frame}{History}

\begin{itemize}
\item Discuss the history of the problem.
\item Describe context for the problem.
\item Outline prior work on the problem.
\end{itemize}

\end{frame}

\section{Representation}

\begin{frame}{Representation}

Represent the problem in symbolic, graphic, or numeric format.

\bigskip

Mathematical formulas may be typeset:
\[ \int_0^\frac{\pi}{2}\frac{1+\cos 2x}{2} \: dx \]

\end{frame}

\section{Methods}

\begin{frame}{Methods and Tools}

Discuss technical methods or tools required to formulate and solve the problem mathematically.

\bigskip

\begin{theorem}
If $f$ is continuous on $[a,b],$ then
$$\int_{a}^{b} f(x) \: dx = F(b) - F(a)$$
where $F$ is any antiderivative of $f,$ that is, a function such that $F' = f.$
\end{theorem}

\end{frame}

\section{Solution}

\begin{frame}{Solution of the Problem}

Present a solution of the problem, perhaps for a simple case, and indicate how the solution may be achieved in other cases.

\bigskip

\begin{example}
\[ \int_0^\frac{\pi}{2}\frac{1+\cos 2x}{2} \: dx=\frac{\pi}{4}  \]
\end{example}

\end{frame}

\section{Conclusion}

\begin{frame}{Conclusion}

Summarize the information presented in the talk.

\bigskip

\begin{itemize}
\item Problem statement
\item Relevance
\item Mathematical tools
\item Solution
\end{itemize}

\end{frame}

\section{References}

\begin{frame}{References}

\begin{thebibliography}{99}
\bibitem{Boas} R. P. Boas,  Can we make mathematics intelligible?  \textit{Amer. Math. Monthly}, \textbf{88} (1981), 727--731.

\bibitem{Page} M. E. Page,  A Brief Citation Guide for Internet Sources in History and the Humanities (Version 2.1),
\begin{url}http://h-net.msu.edu/$\sim$africa/citation.html\end{url}.

\end{thebibliography}

\end{frame}

\end{document}

>>>>

\documentclass{beamer}
\usepackage{amsmath,amsfonts}	% use math symbols
\usepackage{graphicx} % insert images
\usepackage{url} % insert urls
\usepackage{textpos} % package for the positioning
\usepackage{caption}
\usepackage{subcaption}
\usepackage{mathtools}
\usetheme{Ilmenau} % change theme
%% NEW COMMANDS
\def\ci{\perp\!\!\!\perp}
\def\dep{\perp\!\!\!\perp\!\!\!\!\!\!\!/\,\,\,\,}
\begin{document}

%% MAKE TITLE
\title{Learning Causation from Data}
\subtitle{Fundamental of Casual Inference and its Applications}
\author{Huang Xiao} %\\ \small{xiaohu@in.tum.de} }%\thanks{Funded by ARAMiS project}}
%\and Author2 \inst{1} }
%\and Author3 \inst{2}}
\institute[Technische Universit\"at M\"unchen]
{ 
	%\inst{1} 
	Chair of IT Security (I20) \\Department of Informatics \\Technische Universit\"at M\"unchen 
	%\and
	%\inst{2} Cloud Service Lab\\Fraunhofer AISEC
}

@@

%% START PRESENTATION
\input{sections/fundamental}
\input{sections/causalBN}
\input{sections/copulaLearn}
\input{sections/experiments}

\section{Conclusion}
\begin{frame}{Wrap up}
\begin{itemize}
\item Causal Inference is a powerful tool for reasoning and predicting
\item Graphical model has a strong representation on causal conditions
\item Learning causation from data observation is totally feasible
\item Without additional information, only equivalence class can be achieved
\item Copula functions enable efficient parameter estimation
\item PICM based structure learning is also efficient
\item Promising experimental results show the interestingness of causal inference
\end{itemize}
\end{frame}

\section{References}
\begin{frame}{References}
\begin{thebibliography}{99}
\bibitem{Boas} J. Pearls,  Causality: Models, Reasoning, and Inference  \textit{Cambridge University Press}, \textbf{2000}.
\bibitem{Page} Joe Whittaker., Graphical Models in Applied Multivariate Statistics, \textit{John Wiley \& Sons}, Ltd, \textbf{2008}.
\bibitem{Page} Roger B. Nelsen, An Introduction to Copulas, \textit{Springer Science+Business Media}, Inc, \textbf{2006}.
\bibitem{Page} Jonas Peters, et al., Identifying Cause and Effect on Discrete Data using Additive Noise Models, \textit{JMLR}, Proceedings Track 9: 597-604, \textbf{2010}.
\end{thebibliography}
\end{frame}
\end{document}
